The Categorical Structure of Profunctor Optics

### 2017-11-08 paper.tex
Typoes etc

\begin{definition}
Suppose $\C$ has finite products, and let $\C$ act on itself via the bifunctor $\times : \C \times \C \to \C$. The \emph{category of lenses} $\Lens$ in $\C$ is the category of optics with respect to this action: $\Lens = \Optic_\times$.
\end{definition}

@@

\begin{proposition}
\label{prop-OpticImpliesLensLaws}
Suppose $\langle l, r \rangle : S \hto A$ is a pre-lens such that $rl = \id_S$ and $\langle lr \rangle = \langle \phi \times A \rangle$ as an element of $\int^{M \in \M} \C(M \times A, M \times A)$, for some $\phi : M \to M$. Then $p$ obeys the three lens laws.
\end{proposition}
\begin{proof}
By Proposition \ref{prop-onthenose}, we may assume that $lr = \phi \times A$ on the nose.

@@

In favourable conditions, we can upgrade this result to show that the existence of a lens $S \hto A$ implies that $S \cong M \times A$ for some $M$. This is a more general version of the argument given in \cite[Corollary 13]{AlgebrasAndUpdateStrategies} for $\Set$.

@@

\begin{definition}
Suppose $\C$ has finite coproducts, and let $\C$ act on itself via the bifunctor $\sqcup : \C \times \C \to \C$. The \emph{category of prisms} in $\C$ is the category of optics with respect to this action: $\Prism = \Optic_\sqcup$.
\end{definition}

Just as lenses are constructed by two concrete maps $\fget : S \to A$ and $\fput : S \times A \to S$, prisms are constructed by two maps $\freview : A \to S$ and $\fmatching : S \to S \sqcup A$. These names are taken from the Haskell \texttt{lens} library.
\begin{align*}
\PreOptic_\sqcup(S, A) &= \int^{M \in \M} \C(S, M \sqcup A) \times \C(M \sqcup A, S) \\
&\cong \int^{M \in \M} \C(S, M \sqcup A) \times \C(M, S) \times \C(A, S) && \text{(universal property of coproduct)} \\
&\cong \C(S, S \sqcup A) \times \C(A, S) && \text{(co-Yoneda)}
\end{align*}
So, if we are given a pre-prism $\langle l, r \rangle : (S, S) \hto (A, A)$ then associated $\freview$ and $\fmatching$ morphisms are given by $\freview = r \inr$ and $\fmatching = (r\inl \sqcup A)l$

@@

\begin{proposition}
When $\C = \Set$\todo{, and probably more generally,} the third law is implied by the other two.
\end{proposition}
\begin{proof}
The key is that in $\Set$, for any map $f : X \to Y$, the set $Y$ is equal to the union of $\im f$ and its complement. The first law implies that $\freview$ is injective, so $S \cong C \sqcup A$ for some complement $C$. Identifying $A$ with its image in $S$, the second law implies that if $a\in A \subset S$ then $\fmatching(a) = \inr(a)$ and if $c\in C \subset S$ then $\fmatching(c) = \inl(c)$. The third law then follows easily by checking both cases.
\end{proof}

>>>>

\begin{definition}
Suppose $\C$ has finite products, and let $\C$ act on itself via the bifunctor $\times : \C \times \C \to \C$. The \emph{category of lenses} $\Lens$ is the category of optics with respect to this action: $\Lens = \Optic_\times$.
\end{definition}

@@

\begin{proposition}
\label{prop-OpticImpliesLensLaws}
Suppose $\langle l, r \rangle : S \hto A$ is a lens, i.e. that that $rl = \id_S$ and $\langle lr \rangle = \langle \phi \times A \rangle$ for some $\phi : M \to M$. Then $p$ obeys the three lens laws.
\end{proposition}
\begin{proof}
By Proposition \ref{prop-onthenose}, we may assume that $lr = \phi \times A$ on the nose.

@@

In favourable conditions, we can upgrade this result to show that the existence of a lens $S \hto A$ implies that $S \cong M \times A$ for some $M$. This is a generalised from the argument given in \cite[Corollary 13]{AlgebrasAndUpdateStrategies} for $\Set$.

@@

\begin{definition}
Suppose $\C$ has finite coproducts, and let $\C$ act on itself via the bifunctor $\sqcup : \C \times \C \to \C$. The \emph{category of prisms} is the category of optics with respect to this action: $\Prism = \Optic_\sqcup$.
\end{definition}

Just as prelenses correspond to a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$, prisms correspond to pairs of maps $\freview : A \to S$ and $\fmatching : S \to S \sqcup A$. These names are taken from the Haskell \texttt{lens} library.
\begin{align*}
\PreOptic_\sqcup(S, A) &= \int^{M \in \M} \C(S, M \sqcup A) \times \C(M \sqcup A, S) \\
&\cong \int^{M \in \M} \C(S, M \sqcup A) \times \C(M, S) \times \C(A, S) && \text{(universal property of coproduct)} \\
&\cong \C(S, S \sqcup A) \times \C(A, S) && \text{(co-Yoneda)}
\end{align*}
So, if we are given a pre-prism $\langle l, r \rangle : (S, S) \hto (A, A)$ then associated $\freview$ and $\fmatching$ morphisms are given by $\freview = r \inr$ and $\fmatching = (r\inl \sqcup A)l$

@@

\begin{proposition}
When $\C = \Set$\todo{, and probably more generally,} the third law is implied by the other two.
\end{proposition}
\begin{proof}
The key is that in $\Set$, for any map $f : X \to Y$, the set $Y$ is equal to the union of $\im f$ and its complement. The first law implies that $\freview$ is injective, so $S \cong C \sqcup A$ for some complement $C$ \todo{name this map and use it properly}. Identifying $A$ with its image in $S$, the second law implies that if $a\in A \subset S$ then $\fmatching(a) = \inr(a)$ and if $c\in C \subset S$ then $\fmatching(c) = \inl(c)$. The third law then follows easily by checking both cases.
\end{proof}
